\subsection{Gradient descent and line search}
\label{subsec:gradient-descent-line-search}

With the local model from \cref{subsec:optimization-euclidean} in place, we can turn derivative information into an algorithm.
The two design questions are the search direction and the step size.
Gradient descent answers the first question by moving opposite the gradient, while fixed-step and line-search rules answer the second.

\subsubsection{Descent methods}
\label{subsec:descent-methods}

With this Euclidean setup in place, we can describe the generic update template behind many first-order algorithms.
Many iterative methods for~\eqref{eq:unconstrained-min} have the form
\begin{equation}
  x^{(k+1)} = x^{(k)} + t^{(k)} \Delta x^{(k)},
  \qquad k = 0,1,2,\dots,
  \label{eq:generic-descent}
\end{equation}
where $t^{(k)} > 0$ is a step size (learning rate) and $\Delta x^{(k)}$ is a descent direction chosen so that taking a small step decreases $f$.

\begin{definition}[Descent direction]
Let $f$ be differentiable at $x$.
A direction $d$ is a descent direction for $f$ at $x$ if $\nabla f(x)^\top d < 0$.
\end{definition}

\begin{lemma}[Why $\nabla f(x)^\top d < 0$ matters]
If $\nabla f(x)^\top d < 0$, then for all sufficiently small $t>0$,
\[
f(x+td) < f(x).
\]
\end{lemma}
\begin{proof}
Consider $\phi(t) = f(x+td)$.
Then $\phi'(0)=\nabla f(x)^\top d < 0$, so $\phi(t) < \phi(0)$ for all small enough $t>0$.
\end{proof}

\begin{algorithm}
  \caption{Generic descent method}
  \label{alg:generic-descent}
  \begin{algorithmic}[1]
    \Require Objective $f$, initial point $x^{(0)}$, rule for choosing a descent direction, rule for choosing a step size, stopping criterion
    \Ensure Final iterate $x$
    \State $k \gets 0$
    \While{stopping criterion not met}
      \State Choose $\Delta x^{(k)}$ such that $\nabla f(x^{(k)})^\top \Delta x^{(k)} < 0$.
      \State Choose a step size $t^{(k)} > 0$.
      \State $x^{(k+1)} \gets x^{(k)} + t^{(k)}\Delta x^{(k)}$.
      \State $k \gets k + 1$
    \EndWhile
    \State \Return $x^{(k)}$
  \end{algorithmic}
\end{algorithm}

\subsubsection{Step-size selection}
\label{subsec:line-search}

The remaining design choice in \cref{alg:generic-descent} is the step size.
There are two standard regimes.
If a smoothness bound is available, a fixed step size gives a clean baseline and leads to explicit rates.
If that information is unavailable or too conservative, a line search chooses a step size adaptively from local function values.

\input{chapters/01_optimization/chunks/gd_03_body}

When that scale is unknown, backtracking gives a local replacement.
\input{chapters/01_optimization/chunks/gd_07_body}

\subsubsection{Gradient descent on quadratics}

With step-size rules in hand, quadratics provide the cleanest exact benchmark for how first-order methods behave under conditioning.
When a smoothness scale is available in advance, constant-step gradient descent gives a simple baseline, and for strongly convex quadratics the rate can be computed exactly from the spectrum.
\input{chapters/01_optimization/chunks/gd_05_body}
